---
title: "GenomicSelection"
author: "Camilo E. Sánchez & Danilo Moreta"
date: "Cali, CO | 26 February 2024"
output:
  html_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 5
    highlight: haddock
    number_sections: yes
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: no
      toc_depth: 4
  pdf_document:
    toc: yes
  word_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

# 1: Define inputs

```{r}
dir <- 
  'D:/OneDrive - CGIAR/00_CassavaBioinformaticsPlatform/07_GenomicSelection/03_DaRTSeq/'
geno_file <- 'Report_DCas23-8685_SNP.csv'
pheno_file <- 'GS_2022_first_cohort.xlsx'
```

# 2: Setting working directory

```{r}
getwd()
setwd(dir)
getwd()
```

# 3: Install packages

```{r}
if (!require(data.table)) install.packages(data.table)
if (!require(readxl)) install.packages(readxl)
if (!require(tidyverse)) install.packages(tidyverse)
if (!require(rrBLUP)) install.packages(rrBLUP)
```

# 4: Load libraries

```{r}
library(data.table)
library(readxl)
library(tidyverse)
library(rrBLUP)
```

# 5: Read files

```{r}
# Read geno data
geno <- as.data.frame(t(fread(geno_file))) %>%
  slice(-2, -3, -4, -5, -6, -7, -8, -9, -10, -11, -12, -13, -14, -15, -16) %>%
  select(-1, -2, -3, -4)

# Read pheno data, filter by group2, delete unnecessary columns and rename the accessions
pheno <- read_excel(pheno_file) %>%
  dplyr::filter(group == 'group2') %>%
  select(-1, -2) %>%
  rename(Genotype = geno_accession_name)

# Display the first few rows of each data frame to check
head(geno)
head(pheno)
```

# 6: Modify geno file

```{r}
# Replace a value inside the data frame
geno[1,1] <- "Plates"

# Change column and row names
colnames(geno) <- c(geno[1,])
rownames(geno) <- c(1:nrow(geno))

# Delete first row, filter by plate, delete firts column and chage markers names
geno <- geno %>% slice(-1) %>%
  dplyr::filter(Plates == 'Group2_Plate01' | Plates == 'Group2_Plate02' |
                  Plates == 'Group2_Plate03' | Plates == 'Group2_Plate04') %>%
  select(-1) %>%
  rename(Genotype = MarkerName)
```

# 7: Joint accession names

```{r}
# Extract genotypes names of geno and pheno file
geno_names <- as.data.frame(geno$Genotype)
pheno_names <- as.data.frame(pheno$Genotype)

# Separate the names
geno_names <- geno_names %>% 
  separate(`geno$Genotype`, into = c("Date", "Trial", "Genotype"), sep = "_") %>%
  select(Genotype)

# Separate the names
pheno_names <- pheno_names %>% 
  separate(`pheno$Genotype`, into = c("Date", "Trial", "Genotype"), sep = "_")

pheno_names <- replace(pheno_names, is.na(pheno_names), '-')

pheno_names <- pheno_names %>%
  mutate(Genotype = ifelse(Genotype == '-', paste0(Date), Genotype)) %>%
  select(Genotype)

# Find duplicate values between pheno and geno files
duplicate_names <- intersect(geno_names$Genotype, pheno_names$Genotype)

# Replace genotypes names
geno$Genotype <- geno_names$Genotype
pheno$Genotype <- pheno_names$Genotype

#Filter unused genotypes
# Geno file
geno <- geno %>%
  dplyr::filter(Genotype %in% duplicate_names)

# Pheno file
pheno <- pheno %>%
  dplyr::filter(Genotype %in% duplicate_names)
```

# 8: Messages

```{r}
# Print the number of samples and SNPs in genomic data files
message(paste0("Number of samples in geno file: ", nrow(geno) , "\n",
               "Number of samples in pheno file: ", nrow(pheno), "\n\n",
               "WARNING: The samples in both files must be the same"))

# Print the number of SNPs and traits to analyze
message(paste0("Number of SNPs: ", ncol(geno) - 1 , "\n",
               "Number of traits: ", ncol(pheno) - 1))
```

# 9: Re-formatting SNP values

```{r}
geno[geno == 0] <- -1
geno[geno == 1] <- 0
geno[geno == 2] <- 1
geno[geno == '-'] <- NA
```

# 10: Define function: GS_prediction

```{r}
# Define the function
GS <- function(single_trait, pheno, geno, prop, k, method){
  
    # Starting run time
    start_time <- Sys.time()
    
    # Convert geno file into numeric values
    geno <- sapply(geno[,-1], as.numeric)
    
    # Impute and extract imputed values in a data frame
    geno_imputed <- 
      data.frame(
      A.mat(geno, max.missing = 0.5, impute.method = method, return.imputed = T)$imputed
      )

    # Round to -1, 0 and 1 the imputed values
    geno_imputed <- round(geno_imputed, digits = 0)
    
    # Select the trait
    trait <- pheno %>% select(all_of(single_trait)) %>% drop_na()
    
    # Estimate the proportion of samples to use and subset the data based on this
    prop_num <- round(length(rownames(trait)) * prop)
    geno_sub <- subset(geno_imputed, (rownames(geno_imputed) %in% rownames(trait)))

    # Divide data in training and testing sets
    training_samples <- as.matrix(sample(rownames(trait), prop_num))
    testing_samples <- as.matrix(setdiff(rownames(trait), training_samples))
    
    # Select training data
    pheno_training <- as.matrix(trait[training_samples,])
    geno_training <- as.matrix(geno_sub[training_samples,])
    
    # Select testing data
    pheno_testing <- as.matrix(trait[testing_samples,])
    geno_testing <- as.matrix(geno_sub[testing_samples,])
    
    # Training model
    trained_model <- mixed.solve(y = as.matrix(pheno_training), Z = geno_training)
    
    # Marker effect
    marker_effects <- as.matrix(trained_model$u)
    
    # Baseline, intercept
    BLUE <- as.vector(trained_model$beta)
    
    # Predictions
    predict_training <- as.matrix(geno_training) %*% marker_effects
    predict_testing <- as.matrix(geno_testing) %*% marker_effects
    
    # Predictions + BLUE
    predicted_training_result <- as.vector(predict_training[,1] + BLUE)
    predicted_testing_result <- as.vector(predict_testing[,1] + BLUE)
    
    # Accuracy & correlation
    accuracy_testing <- cor(as.vector(pheno_testing), predicted_testing_result, 
                            use = "complete")
    corr_model <- cor(as.vector(pheno_training), predicted_training_result, 
                      use = "complete")
    
    # Ends running time
    end_time <- Sys.time()
    
    # Calculate the time
    running_time <- end_time - start_time

    return(list(trait = single_trait,
                training_prop = prop,
                testing_prop = 1 - prop,
                cv = k,
                prediction_ability = accuracy_testing, 
                time = running_time))

}
```

# 11: Cross validation for single trait

```{r}
# Define inputs
method <- 'mean'
cv <- 100
prop <- 0.75
single_trait <- 'yield_ha'

# Create an empty data frame to storage results
trait_df <- data.frame(matrix(ncol = 6, nrow = 0))

# Create and set a directory to store the results
setwd(dir)
dir.create(paste0('GS_Results_SingleTrait_', single_trait, '_', cv, '_', prop))
setwd(paste0('GS_Results_SingleTrait_', single_trait, '_', cv, '_', prop))

# Define the time of starting
start_time <- Sys.time()

# Perform the function using CV for a single trait
for (k in 1:cv){
    
    # Perform the function
    result <- GS(single_trait, pheno, geno, prop, k, method)
    
    # Store the results
    trait_df <- rbind(trait_df, data.frame(result))
    
    # Progress bar
    cat('\r', k, ' files processed |', rep('=', k / 10), 
        ifelse(k == nrow(cv), '|\n', '>'), sep = '')
    
}

# Save the results
write.csv(trait_df, paste0(single_trait, '_', cv, '_', prop, '.csv'))

# Calculate ending time
end_time <- Sys.time()

# Print running time
message(paste0('Function ends of running. Time taken: ', (end_time - start_time)))
```

# 12: Cross validation for all traits

```{r}
# Define inputs
method <- 'mean'
cv <- 50
prop <- 0.8

# Create an empty data frame to storage results
trait_df <- data.frame(matrix(ncol = 6, nrow = 0))

# Create and set a directory to store the results
setwd(dir)
dir.create(paste0('GS_Results_AllTraits_', cv, '_', prop))
setwd(paste0('GS_Results_AllTraits_', cv, '_', prop))

# Define the time of starting
start_time <- Sys.time()

# Perform the function complete using all traits and CV
for (single_trait in colnames(pheno[,-1])){
  
  # Informative message
  message(paste0('Working on trait: ', single_trait, '\n'))
    
  # Cross validation for each trait
  for (k in 1:cv){
      
    # Perform the function
    result <- GS(single_trait, pheno, geno, prop, k, method)
    
    # Store the results
    trait_df <- rbind(trait_df, data.frame(result))
      
    # Progress bar
    cat('\r', k, ' files processed |', rep('=', k / 10), 
        ifelse(k == nrow(cv), '|\n', '>'), sep = '')
      
    }
    
  # Informative message
  message(paste0('\n', 'Done on trait: ', single_trait, '\n'))
    
}

# Save the results
write.csv(trait_df, paste0('GS_Results_AllTraits_', cv, '_', prop, '.csv'))

# Calculate ending time
end_time <- Sys.time()

# Print running time
message(paste0('Function ends of running. Time taken: ', (end_time - start_time)))
```

# 13: Cross validation using all traits and different prop values

```{r}
# Define inputs
method <- 'mean'
cv <- 20
props <- c(0.9, 0.8, 0.7, 0.6, 0.5)

# Create an empty data frame to storage results
trait_df <- data.frame(matrix(ncol = 5, nrow = 0))

# Create and set a directory to store the results
setwd(dir)
dir.create(paste0('GS_Results_AllTraits_DiffProp_', cv))
setwd(paste0('GS_Results_AllTraits_DiffProp_', cv))

# Define starting point
start_time <- Sys.time()

# Perform the function complete using all traits, CV and different prop
for (single_trait in colnames(pheno[,-1])){
  
  # Informative message
  message(paste0('Working on trait: ', single_trait, '\n'))
    
  # Iterate over the different prop values
  for (prop in props){
    
    # Informative message
    message(paste0('% of validation set: ', 1 - prop, '\n'))
    
    # Cross validation for each trait
    for (k in 1:cv){
      
      # Perform the function
      result <- GS(single_trait, pheno, geno, prop, k, method)
      
      # Store the results
      trait_df <- rbind(trait_df, data.frame(result))
      
      # Progress bar
      cat('\r', k, ' files processed |', rep('=', k / 10), 
          ifelse(k == nrow(cv), '|\n', '>'), sep = '')
      
    }
    
    # Informative message
    message(paste0('\n', 'Done on validation set of: ', 1 - prop, '%', '\n'))
  
  }
  
  # Informative message
  message(paste0('\n', 'Done on trait: ', single_trait, '\n'))
  
}

# Save the results
write.csv(trait_df, paste0('GS_Results_AllTraits_DiffProp_', cv, '.csv'))

# Calculate ending time
end_time <- Sys.time()

# Print running time
message(paste0('Function ends of runnin. Time taken: ', (end_time - start_time)))
```

# 14: Plots

```{r}
# Write the labels
labels <- c('Height', 'Dry matter gravity', 'Root skin color', 'Height 1st branch',
            'Branch angle', 'Plant type', 'Root shape', 'Carotenoid', 
            'Dry matter yield (ha)', 'Root peduncle', '# of commercial roots', 
            'Branch number', 'Thrips', 'Yield (ha)', 'Root type', 'Root weight (Plot)',
            'Height wt leaf', 'Root length', 'Germinated number (plot)', '% Germination',
             'Root constriction')

# Extract data
traininig_value <- unique(trait_df$training_prop)
CV_value <- max(trait_df$cv)

# Accuracy plot for all traits
pdf(paste0('GS_Plot_AccuracyAbility.pdf'), width = 15, height = 10)

ggplot(trait_df, aes(x = prediction_ability, 
                     y = reorder(trait, prediction_ability, FUN = median),
                     fill = trait, alpha = 0.5)) +
  geom_violin(fill = "gray80", color = "white", width = 1, alpha = 0.5) +
  geom_boxplot(width = 0.3) + 
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 14, color = 'black'),
        axis.text = element_text(size = 14, color = 'black'),
        title = element_text(size = 14, color = 'black')) + 
  labs(x = 'Prediction ability', y = 'Traits', 
       title = paste0('Training proportion: ', traininig_value, ', CV: ', CV_value)) +
  xlim(0, 1) +
  scale_y_discrete(labels = labels) +
  geom_vline(xintercept = 0.50, linetype = 'dotted', color = 'gray40', linewidth = 0.80) + 
  geom_vline(xintercept = 0.30, linetype = 'dotted', color = 'gray40', linewidth = 0.80)

dev.off()
```

# 15: Define function: Genomic Estimated Breeding Values (GEBVs)

```{r}
# Define function
GEBVs <- function(single_trait, pheno, geno, testing_names){
  
  # Impute and extract imputed values in a data frame
  Imputed <- 
    data.frame(A.mat((sapply(geno[,-1], as.numeric)), max.missing = 0.5, 
                     impute.method = "EM", return.imputed = T)$imputed)

  # Round to -1, 0 and 1 the imputed values
  geno_tr_model <- round(Imputed, digits = 0)

  # Divide data in training and testing
  training_samples <- data.frame(pheno$Genotype) %>%
    sample_frac(prop_num, replace = F) %>%
    rename(Genotype = 1)

  testing_samples <- data.frame(setdiff(pheno$Genotype, training_samples$Genotype)) %>%
    rename(Genotype = 1)

  # Subset geno and pheno file according to training samples 
  geno_tr <- geno %>% dplyr::filter(Genotype %in% training_samples$Genotype)
  pheno_tr <- pheno %>% dplyr::filter(Genotype %in% training_samples$Genotype)
  
  #
  trait <- pheno_tr %>% select(all_of(single_trait))
  trained_model <- mixed.solve(y = as.matrix(trait), Z = geno_tr_model)
    
  #
  marker_effects <- as.matrix(trained_model$u)
  BLUE <- as.vector(trained_model$beta)
  markers_model <- colnames(geno_tr_model)
    
  # 
  geno_test <- geno %>% 
    dplyr::filter(Genotype %in% testing_samples$Genotype) %>%
    select(-1) %>%
    select(one_of(markers_model))
    
  #
  geno_test <- sapply(as.matrix(geno_test), as.numeric)
    
  #
  predict_gs <- geno_test %*% marker_effects
  predict_gs_result <- data.frame(predicted_value = (predict_gs[,1] + BLUE))

  #
  message(paste0("Saving GEBVs results of: ", single_trait))
  write.csv(predict_gs_result, paste0(single_trait, '_GEBV.csv'))
  return(list('trait' = single_trait,
              'BLUE' = BLUE,
              'mean' = mean(predict_gs_result[,1])))

}
```

# 16: Estimate GEBVs DEFINIR BIEN VOLVER A REVISAR

```{r}
# Define inputs
single_trait <- 'yield_ha'
prop_num <- 0.8

# Create and set a directory to store the results
setwd(dir)
dir.create(paste0('GS_GEBVs_SingleTrait_', single_trait))
setwd(paste0('GS_GEBVs_SingleTrait_', single_trait))

# Define starting point
start_time <- Sys.time()

# GEBVs for all traits
for (single_trait in colnames(pheno)[-1]){
    
    #
    message(paste0('Trait: ', single_trait))

    #
    blup_table <- data.frame(matrix(ncol = 2, nrow = 0))

    # 
    result <- gs_prediction(single_trait, pheno, geno, testing_names)
    blup_table <- rbind(blup_table, data.frame(result))
    
    #
    assign(paste0(single_trait, "_result" ), result)
    save(result, file = paste0(single_trait, "_result.RData"))

}

# Save the results
write.csv(blup_table, paste0('Results_BLUE_MEAN.csv'))

# Calculate ending time
end_time <- Sys.time()

# Print running time
message(paste0('Function ends of running. Time taken: ', (end_time - start_time)))

```
